The effects of drought on land cover change and vegetation productivity in continental Chile

### 2024-04-11 article/paper_drought_Chile2023.qmd
last update antes egu2024

title: The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

@@

          name: Universidad Mayor
          department: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad
          #address: Street Address
          city: Santiago, Chile
          #state: State
          postal-code: 7500994

@@

    note: This is the first author footnote.

@@

          name: University of Twente
          department: Faculty of Geo-Information Science and Earth Observation (ITC)
          #address: Street Address

@@

@@

@@

          name: Universidad San Sebastian
          department: 
          #address: Street Address
          city:  

@@

@@

@@

 The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a megadrought. The aim of our study is to evaluate the land cover change over continental Chile and to examine how this is connected to drought indices of water supply, atmospheric evaporative demand (AED), soil moisture, and their effects on vegetation productivity. The drought indices were derived using monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were utilized to obtain information on annual land cover and monthly vegetation productivity. We analyzed short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales of drought. Our results showed that land cover change was highest in the south-central part of the country, reaching changes as high as 36% in the surface type. The water demand has increased for the whole country, with a major increase in the north. The AED and soil moisture evidence a decreasing trend, which decreases at longer time scales and from north to south. The extreme south part of the country shows an increase in supply. Vegetation productivity has a negative trend in the north-central region for all land cover types. On the other hand, forests seem to be the most resistant type to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are the ones that consider soil moisture for twelve months and the combined effect of precipitation and AED for 24 and 12 months. The results indicate that the north-central region is the most sensitive to water supply deficits lasting longer than a year. 
keywords: 
  - drought
  - land cover change
  - vegetation productivity
  - satellite
date: last-modified
bibliography: references.bib
keep-tex: false
include-before-body:
  - text: |
      \captionsetup{justification=raggedright,singlelinecheck=false}
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
  docx:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}

@@

Drought is often classified as meteorological when there is a decrease in precipitation below the mean average of several years (more than 30 years), hydrological when these anomalies last for long periods (months to years) and affect water systems, and agricultural when the deficit impacts plant health anomalies and leads to decreased productivity [@Wilhite1985]. However, it is important to note that drought is also influenced by human activities, which were not considered in the definitions. Thus, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans’ decisions and activities that drives the anthropogenic drought. Simultaneously, drought leads to heightened tree mortality and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017]. Even though many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. Then, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems." In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on water security. [@Bakker2012]

Human-induced greenhouse gas emissions have increased the frequency and/or intensity of drought as a result of global warming, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, suggesting that drought has not exhibited a significant trend over the past forty years. [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the meteorological drought trend on a global scale, finding that only in a few regions has there been an increase in the severity of drought. Moreover, they attribute the increase in droughts over the past forty years solely to an increase in atmospheric evaporative demand (AED), which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that "the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change". Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years. Further, the @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments are needed that relate meteorological and soil moisture variables to their effects on vegetation.

From 1960 to 2019, land use change has impacted around one-third of the Earth's surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021; @Song2018] and regionally [@Chamling2020; @Homer2020; @Yang2021]. Some others seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on the interaction between drought and land cover change [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 conducted a worldwide investigation utilizing net primary production to examine the spatial and temporal variations in vegetation productivity at global level. The study aimed to assess the influence of drought by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline in vegetation productivity, while water availability accounts for 55% of the variation. @Chen2022 studied the trend of vegetation greenness and productivity and its relation to meteorological drought (SPEI of twelve months in December) and soil moisture at the global level. The results showed lower correlations (<0.2) for both variables. @Akinyemi2021 evaluates drought trends and land cover change using vegetation indices in Botwsana in a semi-arid climate. These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) over a single time period of 12 months. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own. Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological and soil moisture droughts? ii) How is the drought impacting land cover changes? And iii) How do they behave in humid and arid climatic zones regarding drought? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017]. Nowadays, there is an increase in attention toward using AED separately to monitor droughts [@Vicente-Serrano2020]. One reason is due to its attribution to increasing flash droughts in water-limited regions [@Noguera2022]. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporated the temperature effect by subtracting AED from precipitation. SPEI allows for analysis of the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2021]. @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from the earth's water balance (e.g., precipitation, AED, soil moisture).  The variation in climate variables impacts vegetation development, and unfavorable conditions such as low precipitation and high temperatures usually generate a decrease in vegetation productivity. To monitor the response of vegetation, the common practice is to use satellite data. The Normalized Difference Vegetation Index (NDVI) has been widely used as a proxy for biomass production [@Camps-Valls2021;@Paruelo2016;@Helman2014]. For Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in response to drought.  Using this information, we can advance our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega drought. [@Garreaud2017], which has impacted the Chilean ecosystem. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016]. We found one study regarding land cover and drought in Chile. The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude. @Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results. For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI. The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. The association between drought and the environment in Chile is not well comprehended. Hence, it is imperative to acquire a more profound comprehension of the manner in which climatic and soil moisture droughts influence environmental dynamics, in order to make well-informed decisions on adaptation strategies.

<!-- A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture. Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially. In order to do this, we use reanalysis data from ERA5-Land [@MunozSabater2021] to create drought indices that consider AED, precipitation, and soil moisture over a range of time periods, from the short to the long term. Also, we use vegetation spectral information and annual land cover change from the Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity.  -->

@@

Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity [@Luebert2022] (@fig-studyArea c). The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009]. "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

To analyze land cover change, we use the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 collection 6.1 from MODIS. The MCD12Q1 has a yearly frequency from 2001 to 2022 and defines 17 classes. To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. MOD13A3 provides vegetation indices at 1km of spatial resolution and monthly frequency. The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.

```{=latex}
\begin{table}[!ht]
\caption{Description of the satellite and reanalysis data used}
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}
```

For soil moisture, water supply, and water demand variables, we used ERA5L (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3). Table \ref{tab-desEOD} shows a summary of the data and its main characteristics.

## Short- to long-term drought

@@

In order to compute the drought indices that use water demand, it is necessary to first calculate the AED. To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:

@@

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m. For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel. We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods (e.g., Penman-Monteith) when the access to climatic variables is limited [@Vicente-Serrano2014].

@@

$$DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}

>>>>

title: The effects of drought on land cover change and vegetation productivity in continental Chile

@@

          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
          city: Santiago
          country: 7500994, Chile
        - id: odes
          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
          city: Santiago
          country: 7500994, Chile

@@

    #note: This is the first author footnote.

@@

          name: Faculty of Geo-Information Science and Earth, University of Twente

@@

        - id: agrouc
          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
          city: Santiago
          country: Chile
        - id: instuc
          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile
        - id: cambioglobal
          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile, Santiago, Chile.

@@

        - id: umayor-io
          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
          city: Santiago
          country: 7500994, Chile

@@

          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
          city: Santiago
          country: Chile
        - id: capes
          name: Center of Applied Ecology and Sustainability (CAPES)
          city: Santiago
          country: Chile

@@

        - id: agrouc

@@

        - id: gema
          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750, Santiago, Chile
        - id: do
          name: Data Observatory Foundation, Santiago, Chile

@@

 Chile has experienced a persistent decrease in water supply, which impacts the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a mega-drought. There are few studies about the relationship between drought and ecosystem changes that can help with a better understanding of ecological drought. The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile. To assess drought, we used drought indices for atmospheric evaporative demand (AED), water supply, and soil moisture from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales. We derived the drought indices using monthly ERA5-Land reanalysis data from 1981 to 2023. We used the Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity. Our results showed that except for the Austral part, Chile has a temporal decreasing trend in water supply, and across the whole country, there is an increase in AED. These trends become stronger over longer time scales. We found a negative trend in vegetation productivity in the north-central area, which is more prominent for shrubland and savanna as compared to croplands and forests. The anomaly in soil moisture over the past 12 months (SSI-12) is the most important variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years (SPI-12 and SPI-24). The variable importance obtained by random forest models indicates that drought is explaining about 20–30% of the change in land cover surface across Chile for forest, grassland, shrubland, and savanna but has no relation to the changes in croplands. The increase in AED is the main variable associated with the change in land cover, followed by a reduction in precipitation and soil moisture. Our findings provide insightful information that could assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought. Also, this study could contribute to a better comprehension of ecological drought.

keywords: 
  - drought
  - land cover change
  - vegetation productivity
  - ecosystem
date: last-modified
bibliography: references.bib
keep-tex: false
include-before-body:
  - text: |
      \captionsetup{justification=raggedright,singlelinecheck=false}
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
  elsevier-docx:
    keep-tex: true
    number-sections: true
    journal:
      name: Journal Name
      formatting: preprint
      layout: onecolumn
      model: 3p
      cite-style: authoryear
      natbiboptions: numbers
    include-in-header: 
      text: \usepackage{lineno}\linenumbers
            \usepackage{multirow}
            \usepackage{lscape}
            \newcommand{\blandscape}{\begin{landscape}}
            \newcommand{\elandscape}{\end{landscape}}
---

@@

Drought is often classified as 1) meteorological when precipitation in a specific period remains below the mean precipitation experienced in the same period during multiple years (more than 30 years usually), 2) hydrological when these anomalies last for long periods (months to years) and affect water systems, and 3) agricultural when the deficit negatively impacts plant health and leads to decreased productivity of crops or pastures [@Wilhite1985]. However, because drought is also influenced by human activities,  @Loon2016 and @AghaKouchak2021 expanded the drought definition for the Anthropocene, indicating that the feedback of human decisions and activities should also be considered (i.e.,  anthropogenic drought). Droughts lead to increased tree mortality [@Cheng2024] and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017]. Even though many ecological studies have at times mistakenly considered "dry" conditions as "drought" [@Slette2019]. Ecological drought can be defined as _"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"_ [@Crausbay2017]. In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on future water security [@Bakker2012].

Global warming, as a result of human-induced greenhouse gas emissions, has increased the frequency and intensity of drought, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, with some suggesting that drought has not exhibited a significant trend over the past forty years [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the trend in meteorological drought on a global scale, finding that only in a few regions an increase in the severity of drought was observed. Moreover, they attributed this increase solely to an increase in atmospheric evaporative demand (AED) due to higher temperatures, which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change, led by an increase in agricultural land"_. Similarly, @Kogan2020 analyzed the drought trend using remotely-sensed vegetation health indicators, finding that for the globe and main grain-producing countries, drought has not expanded or intensified during the past 38 years. Nonetheless, @IPCC2021 suggests that there is a medium to high degree of confidence that rising temperatures will increase the extent, frequency, and severity of agricultural and ecological droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments that correlate meteorological and soil moisture variables to their effects on vegetation are much needed.

From 1960 to 2019, land use change has impacted around one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021;@Song2018] and regionally [@Chamling2020;@Homer2020;@Yang2021;@Schulz2010;@Echeverria2012]. Some seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on drought and its relation to land cover change and vegetation productivity [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 utilized net primary productivity to examine the spatial and temporal variations in vegetation productivity at global level and assess to what extent drought influenced this variability by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline and accounts for 55% of the variability in vegetation productivity. @Chen2022 instead found poor correlations (r<0.2) between the vegetation productivity trends against meteorological drought (SPEI of twelve months in December) and soil moisture at the global level.  These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) obtained for 12 month periods. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow to know the contribution of each variable on its own. To better understand these contributions on land cover change and vegetation productivity the following questions may be asked: i) how do land cover and vegetation productivity respond to short- to long-term meteorological and soil moisture droughts? And ii) How is this response different between humid and arid climatic zones? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

To address the previous questions over extensive regions, we can utilize gridded data on water availability, vegetation conditions, and the respective drought indices. For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporates the temperature effect by subtracting AED from precipitation. SPEI allows for analyzing the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2024]. Recently, there has been more interest in using AED to track droughts separately to better disengage the effects of precipitation from temperature-dependent effects [@Vicente-Serrano2020]. One of the reasons is that AED is linked to more flash droughts in limited water regions [@Noguera2022].  @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, many drought indices exist that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from Earth’s water balance (e.g., precipitation, AED, soil moisture). Climatic variability impacts vegetation development, with unfavorable conditions such as low precipitation and high temperatures usually promote a decrease in plant productivity. To monitor the response of vegetation for large areas, the common practice is to use satellite data. For example, the Normalized Difference Vegetation Index (NDVI) derived from frequent satellite observations of red and near infrared spectral reflectance, has been widely used as a proxy for biomass production [@Camps-Valls2021; @Paruelo2016;@Helman2014]. For Chile's cultivated land, @Zambrano2018 used the zcNDVI for assessing seasonal biomass production in response to drought. Comparing the various meteo-related and vegetation-based drought indices, we can further our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega-drought [@Garreaud2017], which has impacted the Chilean ecosystem and consequently makes it highly vulnerable to climate change [@Barria2021;@AlvarezGarreton2021]. This mega-drought was defined by the annual time series of the Standardized Precipitation Index (SPI) at a time scale of twelve months at the end of each year (December) when having values below one standard deviation.  Some studies have addressed how this drought affects single ecosystems in terms of forest growth [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016].  The term "mega-drought" is used in Chile to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. In Chile, the relationship between drought and the environment remains poorly understood. Hence, we aim to contribute to understanding how climatic and soil moisture droughts influence ecosystem dynamics in order to provide useful information that helps for a better understanding of ecological droughts and, at the same time, helps to make well-informed decisions on adaptation strategies.

@@

Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] ([@fig-studyArea]a), which determines its great ecosystem diversity [@Luebert2022] ([@fig-studyArea]c). The Andes Mountains are a main factor in climate variation [@Garreaud2009]. For an aggregated overview of the results of the study, we used the five Chilean macrozones: "Norte Grande," "Norte Chico," "Centro," "Sur," and "Asutral". "Norte Grande" (17°34’–25°42’S) and "Norte Chico" (25°42'-32°8'S) predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland. In the macrozones "Centro" (32°08'-36°12'S) and the northern half of "Sur," (36°12'-43°48'S) the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral" (43°48'-56°00'S) Those zones have a large areal extent of forest and grassland. The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

To analyze land cover change, we used the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for each year from 2001 to 2022 and defines 17 classes (see Table S1). To maintain our focus on a large scale and follow the FAO classification [@FAO2022], we considered native and planted forests as "forests", which represent ecosystems dominated by larger trees. To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 Collection 6.1 from MODIS [@Didan2015]. MOD13A3 provides vegetation indices with 1km spatial resolution and monthly frequency. The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.

```{=latex}
\begin{table}[!ht]
\caption{Description of the satellite and reanalysis data used}
\label{tab-desEOD}
\scriptsize
\centering
\begin{tabular}{p{0.1\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}
```

For soil moisture, water supply, and water demand variables, we used ERA5L (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3). \ref{tab-desEOD} shows a summary of the data and its main characteristics.

## Short- to long-term drought trends

@@

To compute the drought indices that use water demand, it is necessary to first calculate the AED. To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:

@@

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m. For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel as follow:

$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}
where

$Ra$: extraterrestrial radiation $[MJ\, m^{-2} day-1]$,  
$G_{sc}$: solar constant = 0.0820 $[MJ\,m^{-2} min^{-1}]$,    
$d_r$: inverse relative distance Earth-Sun,  
$\omega_s$ sunset hour angle  $[rad]$,  
$\phi$: latitude $[rad]$,  
$\delta$: solar declination $[rad]$.

We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods (e.g., Penman-Monteith) when the access to climatic variables is limited [@Vicente-Serrano2014].

@@

$$DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}

$DI$ is referring to the drought index calculated for the variable $V$ (i.e., SPI, SPEI, EDDI, SSI, and zcNDVI). The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$. For $P(A^s_i) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A^s_i))}$ , and for $P(A^s_i) > 0.5$, replace $P(A^s_i)$ with $1-P(A^s_i)$ and reverse the sign of $DI(A^s_i)$.

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought.
